Learning with misspecified models: Overconfidence and Stereotypes

### 2023-09-07 Paper/draft.tex
editing text

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  12pt,
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Learning from Data Through Models},
  pdfauthor={Jimena Galindo},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=2.5cm]{geometry}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}

@@

\title{Learning from Data Through Models}
\author{Jimena Galindo}
\date{September 05, 2023}

@@

An agent has a \emph{misspecified} belief if the prior assigns
probability zero to their true type. Furthermore, the agent is
\emph{dogmatic} if he holds a degenerate belief that places probability
one on being a particular type, \(\hat{\theta}\). A dogmatic agent can
be dogmatic and misspecified. That means that
\(\hat{\theta} \neq \theta^*\) and \(p_0(\hat{\theta}) = 1\).

The agent chooses an action \(a\in A\) and observes a noisy outcome
\(h\in H\). The outcome is a function of the agent's type, the state,
and the action. In particular
\(h = h(\theta^*, \omega^*, a) + \varepsilon\) with \(h(\cdot)\)
increasing in both \(\theta^*\) and \(\omega^*\); and such that
conditional on a pair of parameters \((\theta, \omega)\), there is a
unique optimal action. \(\varepsilon\sim N(0, \sigma)\) is noise in the
output.

After observing the outcome, the agent updates his beliefs about
\(\theta\) and \(\omega\) using some algorithm and moves on to the next
period. He repeats this process infinitely many times. I make the
simplifying assumption that the agent is myopic and chooses the action
that maximizes the payoff in each period. This assumption is crucial for
the results as the agent will not trade off the flow payoff for learning
the truth as a forward-looking agent would.

Within this framework, I consider two nested theories of belief
updating. The first one is a dogmatic modeler from \citet{Heidhues2018}.
The second one is a switcher, as in \citet{Ba2023}. The dogmatic modeler
can be seen as a switcher with an infinitely sticky initial belief and
this is the sense in which the two are nested.

A key notion in this setting is that of a self-defeating
equilibrium\footnote{This notion is an adaptation of the Berk-Nash equilibrium in 
@Esponda2016}. A \emph{self-defeating equilibrium} is a belief and
action pair such that the agent's belief about their type is
misspecified and the outcome generated by the action is consistent with
the misspecified belief. The average outcome under the true type and the
true state equals the average output the agent expects under the
misspecified belief. In addition, the agent's belief is said to be
\emph{stable} when this happens.

@@

A dogmatic agent does not update their beliefs about \(\theta\); instead
he holds a degenerate belief that places probability one on being a
particular type, \(\hat{\theta}\), which is potentially not his true
type. In this case, no matter how much information he gathers against
being of type \(\hat{\theta}\), he will not update his beliefs. Any
discrepancies between the observed outcomes and his believed type are
incorporated using the Bayes rule to update their beliefs about
\(\omega\).

A key difference between the dogmatic agent and a fully Bayesian agent
is that the dogmatic agent does not aggregate across types when updating
their beliefs about \(\omega\). This means the dogmatic agent will never
learn their true type if they are misspecified.

\citet{Heidhues2018} show that, under certain assumptions on the
per-period utility,\footnote{The assumptions are that 
$u$ is twice continuously differentiable with: (i)$u_{ee}<0$ and $u_e(\underline{e} \theta, \omega)>0>u_e(\Bar{e}, \theta, \omega)$, 
(ii) $u_{\theta}, u_{\omega}>0$ and (iii) $u_{e\theta}<0$ and $u_{e\omega}>0$. The direction of the derivatives is a normalization
and the results would hold even when the signs are reversed.} about the
outcome \(h\), a dogmatic modeler will inevitably fall into a
self-defeating equilibrium where the outcomes they observe reinforce
their belief on \(\omega\) in such a way that as \(t\to\infty\) the
agent will be sure that the state is some \(\omega^{'}\) consistent with
their believed type and the observed data.

The mechanism by which the dogmatic agent falls into the self-defeating
equilibrium is the following: Suppose the agent holds the misspecified
belief that they are type \(\hat{\theta}>\theta^*\). For any prior over
\(\omega\), the agent will be disappointed by the outcome; he expected a
gain of \(h(\hat{\theta}, \mathbb{E}(\omega), a)\) but instead observes
\(h(\hat{\theta}, \omega^*, a)\). There are two possible sources for the
disappointment. The first is that the realized state is lower than the
expected state. The second is that the agent is of type \(\theta^*\) and
thus for all possible states, his gain will be lower than what he
expected. Because the agent is dogmatic, he will not update his beliefs
about \(\theta\) and thus will attribute the disappointment to the state
being lower than expected. \citet{Heidhues2018} give sufficient
conditions for the agent to fall into a self-confirming equilibrium
where the agent will continue to choose an action that reinforces an
incorrect belief about the state. This force is illustrated in Example
1.

\textbf{Example 1: } Set \(A = \Omega\) and \(H = [0, \infty)\) and
consider a student with intrinsic ability \(\theta^*\geq 0\) who faces a
grading procedure \(\omega^*\) that is unknown to them. However, they
know that a higher \(\omega^*\) is more likely to yield a higher grade.
In particular, assume the grade is given by \((\theta^*+a)\omega^*\).

The student must choose an effort level \(a\), which determines their
grade. For whatever the chosen effort level, the agent must pay a cost
\(c(a) = \frac{1}{2}a^2\). And he repeats this process for infinitely
many periods. Assume also that the student's prior is such that
\(mathbb{E}[\omega]= \omega^*\). Therefore, the student's payoff in
period \(t\) is given by

@@

Under this specification, the myopic optimal effort level is
\(a_t^* = \omega^*\). Nonetheless, because the agent does not know
\(\omega^*\), he will choose \(a_t = \mathbb{E}_t(\omega)\) where the
expectation is taken with respect to the agent's belief at the beginning
of period \(t\). If he does not revise his effort choice for multiple
periods, he will observe an average grade of
\((\theta^*+a_t^*)\omega^* - \frac{1}{2}a_t^{*2}\) but he was expecting
an average grade of
\((\hat{\theta}\theta+a_t^*)\omega^* - \frac{1}{2}a_t^{*2}\)

\hypertarget{the-bayesian-benchmark}{%
\subsection{The Bayesian Benchmark}\label{the-bayesian-benchmark}}

A Bayesian agent simultaneously updates their beliefs about \(\theta\)
and \(\omega\) by using Bayes' rule. The posterior odds at period t
about \(\theta\) after observing an outcome are given by:
\begin{equation}
\frac{p_{t}[\theta_H|\text{outcome}]}{p_{t}[\theta_M|\text{outcome}]} = 
      \frac{p[\text{outcome}|\theta_H]p_{t-1}[\theta_H]}{p[\text{outcome}|\theta_M]p_{t-1}[\theta_M]}
\end{equation} and \begin{equation}
\frac{p_{t}[\theta_M|\text{outcome}]}{p_{t}[\theta_L|\text{outcome}]} = 
      \frac{p[\text{outcome}|\theta_M]p_{t-1}[\theta_M]}{p[\text{outcome}|\theta_L]p_{t-1}[\theta_L]}
\end{equation}

Bayesian agents always choose the effort level that maximizes their flow
payoff by taking expectations over their prior beliefs about \(\theta\)
and \(\omega\). Since agents are myopic, even though all the parameters
could be identified with enough variation in choices,

@@

An agent is a \emph{switcher} if they behave as a dogmatic but is
willing to entertain the possibility that they are of a different type.
In particular, when they start off as a misspecified dogmatic, they are
willing to switch to a different dogmatic belief if the data is
convincing enough.

@@

\[\frac{p[h^t|\tilde{\theta}]}{p[h^t|\hat{\theta}]} > \alpha\]

Where \(h^t\) is the history of outcomes up to time \(t\) and
\(\alpha \geq 1\) is a threshold that determines how convincing the data
needs to be for the agent to change their belief. Notice that if
\(\alpha \to \infty\), we get the dogmatic agent. In this sense, the
switcher is a generalization of the dogmatic type, just as the
self-attribution agent is a generalization of the Bayesian agent.

By allowing the agent to keep track of the likelihoods and switching to
an alternative type, the switcher can avoid the self-confirming
equilibrium the dogmatic agent falls into. However, if the prior belief
on \(theta\) is sufficiently tight around a self-confirming equilibrium,
the switcher might look identical to the dogmatic even in a case where
\(\alpha\) is not too large.

EXPLAIN HOW THEY MIGHT GET OUT OF THE TRAP.

\hypertarget{the-bayesian-agent}{%
\subsection{The Bayesian Agent}\label{the-bayesian-agent}}

a fully Bayesian agent might not learn their true type. This happens
because they do not internalize the tradeoff between flow payoff and
learning and thus might not experiment enough to learn their type. An
alternative to this approach is given by \citet{Hestermann2021} and is
discussed with the results.

\hypertarget{the-biased-agent}{%
\subsection{The Biased Agent}\label{the-biased-agent}}

An agent that updates their beliefs with self-attribution bias will
update their beliefs about the state \(\omega\) and their type
\(\theta\) by over-attributing successes to a high value of \(\theta\)
and under-estimating the role of higher \(\omega\). Similarly, they will
attribute failure to a low state more than an unbiased agent would. To
model the self-serving attribution bias, I take the approach of
\citet{benjamin2019}, where a generalization of the Bayes rule above
gives the update.

\begin{equation}
\frac{p_{t}[\theta_H|\text{outcome}]}{p_{t}[\theta_M|\text{outcome}]} = 
      \left(\frac{p[\text{outcome}|\theta_H]}{p[\text{outcome}|\theta_M]}\right)^{c_s^{\theta}\mathbb{I}\{\text{success}\}+c_f^{\theta}\mathbb{I}\{\text{failure}\}}\frac{p_{t-1}[\theta_H]}{p_{t-1}[\theta_M]}
\end{equation} and \begin{equation}
\frac{p_{t}[\theta_M|\text{outcome}]}{p_{t}[\theta_L|\text{outcome}]} = 
      \left(\frac{p[\text{outcome}|\theta_M]}{p[\text{outcome}|\theta_L]}\right)^{c_s^{\theta}\mathbb{I}\{\text{success}\}+c_f^{\theta}\mathbb{I}\{\text{failure}\}}\frac{p_{t-1}[\theta_M]}{p_{t-1}[\theta_L]}
\end{equation}

@@

\hypertarget{a-simple-example}{%
\section{A Simple Example}\label{a-simple-example}}

The agent can be of one of 3 types:
\(\theta \in \{\theta_L, \theta_M, \theta_H\}\) with
\(\theta_H > \theta_M > \theta_L\). They face an unknown exogenous
success rate \(\omega \in \{\omega_L, \omega_M, \omega_H\}\) with
\(\omega_H>\omega_M>\omega_L\). Each of the values of \(\omega\) is
realized with equal probability. The agent knows the distribution of
\(\omega\) but not its realized value. Let the true type be \(\theta^*\)
and let the true state be \(\omega^*\).

The agent holds some prior belief about \(\theta\)
\footnote{which is potentially misspecified as in the dogmatic and switcher cases discussed}
and chooses a binary gamble \(e in \{e_L, e_M, e_H\}\). The agent
observes whether the gamble is a success or a failure and gets a payoff
of \(1\) and if it is a success. They get \(0\) otherwise.

The probability of success is increasing in both \(\theta\) and
\(\omega\) and is fully described in the following table:

@@

Conditional on a type, the agent's flow payoff is maximized by choosing
the gamble that matches the state. For example, if the value of
\(\omega\) is \(\omega_H\), the agent's flow payoff is maximized by
choosing \(e_H\) and if the state is \(\omega_L\) the flow payoff is
maximized by choosing gamble \(e_L\). The agent myopically chooses
gambles every period to maximize the flow payoff for \(T\) periods.

After observing the outcome of each gamble, the agent updates their
beliefs using some algorithm and moves on to the next period.

Notice that both \(\theta\) and \(\omega\) can be identified from the
outcomes if enough variation in the effort choices exists. This can be
seen by confirming there is no pair of \(\theta\) and \(\omega\) such
that the probability of success is the same for all effort choices.
Therefore, by changing the effort choice, the agent can learn both their
type and the state if they observe enough outcomes.

In this example, for an agent with a dogmatic belief about their type, a
self-defeating equilibrium is one in which the agent chooses an effort
level that, under the true \(\theta\), yields a frequency of success
that is consistent with the agent's misspecified belief. That is, the
observed rate of success under the true type and the effort level
chosen, \(P[\text{sucess}|\theta^*, e]\) is the same as the probability
of success under the agent's belief,
\(P[\text{sucess}|\hat{\theta}, e]\).

In the data-generating process described above, there are five such
equilibria. For example, if the agent is of type \(\theta_M\) but
mistakenly believes that he is of type \(\hat{\theta}=\theta_H\) and the
and \(\omega^* = \omega_M\), when the effort chosen is \(e_L\), the
agent will observe a success with 45\% chance. Because the agent
dogmatically believes that their type is high, they will erroneously
conclude that the rate is \(\omega_L\). Under this belief, the optimal
action is \(e_L\) which will continue to generate successes with \(45%
\) probability, further reinforcing the incorrect belief. If the agent
were to learn the true type and rate, they would instead choose \(e_M\),
yielding a success with \(65%
\) probability and strictly higher expected payoffs.

From \citet{Heidhues2018}, we know that if \(\theta^*\) and \(\omega^*\)
are such that there is a self-confirming equilibrium, a dogmatic agent
will always fall into the trap and continue to make sub-optimal choices.
On the other hand, a switcher will be able to learn the true type and
state as long as their prior is diffused enough and their switching
threshold is low enough as in \citet{Ba2023}.

By including self-confirming equilibria, the example captures the forces
from each of the updating mechanisms discussed in the previous section
and allows for direct comparison of all the theories. For realizations
of \((\theta, \omega)\) for which there are self-confirming equilibria,
the dogmatic agent will fall into the trap whereas the switcher will be
able to escape it. Similarly, an agent with self-attribution bias will
update their beliefs differently from an unbiased Bayesian, leading them
to choose different gambles. I exploit such cases in order to test which
model is a better fit for how subjects behave in a laboratory
experiment.

ILLUSTRATE THE DIFFERENT PATHS WITH THE SIMULATION.

\hypertarget{the-dogmatic-modeler-1}{%
\subsection{The Dogmatic Modeler}\label{the-dogmatic-modeler-1}}

A dogmatic modeler updates their beliefs about omega by using bayes
rule. For the example, that means he ends up with the following
posterior odds ratios:

\begin{equation}
\frac{p_{t}[\omega_H|\text{outcome}]}{p_{t}[\omega_M|\text{outcome}]} = 
      \frac{p[\text{outcome}|\omega_H, \hat{\theta}, e]p_{t-1}[\omega_H]}{p[\text{outcome}|\omega_M, \hat{\theta}, e]p_{t-1}[\omega_M]}
\end{equation} and \begin{equation}
\frac{p_{t}[\omega_M|\text{outcome}]}{p_{t}[\omega_L|\text{outcome}]} = 
      \frac{p[\text{outcome}|\omega_M, \hat{\theta}, e]p_{t-1}[\omega_M]}{p[\text{outcome}|\omega_L, \hat{\theta}, e]p_{t-1}[\omega_L]}
\end{equation}

@@

The experiment consisted of 2 treatments: the ego-relevant condition and
the stereotype condition. Subjects participated in only one of them.
Treatments were randomly assigned at the session level. The tasks were
identical across treatments, but in the ego condition the probability of
success was based on the subject's own performance in a quiz, while in
the stereotype condition, the probability od success was based on the
performance of a randomly selected subject from another session.

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  12pt,
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Learning with misspecified models: Overconfidence and Stereotypes},
  pdfauthor={Jimena Galindo},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=2.5cm]{geometry}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}

@@

\title{Learning with misspecified models: Overconfidence and
Stereotypes}
\author{Jimena Galindo}
\date{September 07, 2023}

@@

An agent has a \emph{misspecified} belief if the prior assigns
probability zero to their true type. Furthermore, the agent is
\emph{dogmatic} if he holds a degenerate belief that places probability
one on being a particular type, \(\hat{\theta}\). An agent can be
dogmatic and misspecified; that means that
\(\hat{\theta} \neq \theta^*\) and \(p_0(\hat{\theta}) = 1\).

The agent chooses an action \(a\in A\) and observes a noisy outcome
\(h\in H\). The outcome is a function of the agent's type, the state,
and the action. In particular
\(h = h(\theta^*, \omega^*, a) + \varepsilon\) with \(h(\cdot)\)
increasing in both \(\theta^*\) and \(\omega^*\), and such that
conditional on a pair of parameters \((\theta, \omega)\), there is a
unique optimal action. \(\varepsilon\sim N(0, \sigma)\) is noise in the
output.

After observing the outcome, the agent updates his beliefs about
\(\theta\) and \(\omega\) using some algorithm and moves on to the next
period. He repeats this process infinitely many times. I make the
simplifying assumption that the agent is myopic and chooses the action
that maximizes the payoff in each period. This assumption is symplifies
the analysis and plays a role on whether an agent who updates their
beliefs using bayes rule would learn the truth or not.

A key notion in this setting is that of a self-defeating
equilibrium\footnote{This notion is an adaptation of the Berk-Nash equilibrium in 
@Esponda2016 to this setting with only one agent}. A
\emph{self-defeating equilibrium} is a belief and action pair such that
the agent's belief about their type is misspecified and the outcome
generated by the action is consistent with the misspecified belief. The
average outcome under the true type and the true state equals the
average output the agent expects under the misspecified belief. In
addition, the agent's belief is said to be \emph{stable} when this
happens.

Within this framework, I consider two nested theories of belief
updating. The first one is a dogmatic modeler from \citet{Heidhues2018}.
The second one is a switcher, as in \citet{Ba2023}. The dogmatic modeler
can be seen as a switcher with an infinitely sticky initial belief and
this is the sense in which the two are nested. Both theories produce
different predictions with respect to the equilibrium outcome.

@@

A dogmatic agent does not update their beliefs about \(\theta\); instead
he holds a degenerate belief that places probability one on being a
particular type, \(\hat{\theta}\), which is potentially not his true
type. In this case, no matter how much information he gathers against
being of type \(\hat{\theta}\), he will not update his beliefs. Any
discrepancies between the observed outcomes and his believed type are
incorporated using the Bayes rule to update their beliefs about
\(\omega\). This means the dogmatic agent will never learn their true
type if they startu off as being misspecified.

\citet{Heidhues2018} show that, under certain assumptions on the
per-period utility,\footnote{The assumptions are that 
$u$ is twice continuously differentiable with: (i)$u_{ee}<0$ and $u_e(\underline{e} \theta, \omega)>0>u_e(\Bar{e}, \theta, \omega)$, 
(ii) $u_{\theta}, u_{\omega}>0$ and (iii) $u_{e\theta}<0$ and $u_{e\omega}>0$. The direction of the derivatives is a normalization
and the results would hold even when the signs are reversed.} a dogmatic
modeler will inevitably fall into a self-defeating equilibrium. The
equilibrium will be such that the outcomes they observe reinforce their
belief on \(\omega\) in such a way that as \(t\to\infty\) the agent will
be sure that the state is some \(\omega^{'}\) consistent with their
believed type and the observed data. In other words, they will be in a
self-defeating equilibrium whith a stable belief that places probability
one on the incorrect parameters \((\hat{\theta}, \omega_{\infty})\)

The mechanism by which the dogmatic agent falls into the self-defeating
equilibrium is the following: Suppose the agent holds the misspecified
belief that they are type \(\hat{\theta}>\theta^*\). For any prior over
\(\omega\), the agent will be disappointed by the outcome. He expected a
gain of \(h(\hat{\theta}, \mathbb{E}(\omega), a)\) but instead observes
\(h(\hat{\theta}, \omega^*, a)\). There are two possible sources for the
disappointment, the first is that the realized state is lower than the
expected state. The second source is that the agent is of type
\(\theta^*\) and therefore, for all possible states, his gain will be
lower than what he expected. Because the agent is dogmatic, he will not
update his beliefs about \(\theta\) and as a consequence will attribute
the disappointment to the state being lower than expected. He will
continue to update in ths way until he converges to a belief about
\(\omega\) that is stable. Such a belief will explain the observed
utility perfectly and allow the agent to rationalize his dogmatic belief
about \(\theta\). Under the assumptions of \citet{Heidhues2018}, there
is a unique value of \(\omega\) at which the belief is stable, I will
refer to such value as \(\omega_\infty\). This mechanism is further
illustrated in Example 1.

\textbf{Example 1: } \emph{Set \(A = \Omega\) and \(H = [0, \infty)\)
and consider a student with intrinsic ability \(\theta^*\geq 0\) who
faces a grading procedure \(\omega^*\) that is unknown to them. However,
they know that a higher \(\omega^*\) is more likely to yield a higher
grade. In particular, assume the grade is given by
\((\theta^*+a)\omega^*\).}

\emph{The student must choose an effort level \(a\), which determines
their grade. For whatever the chosen effort level, the agent must pay a
cost \(c(a) = \frac{1}{2}a^2\). And he repeats this process for
infinitely many periods. Assume also that the student's prior is such
that \(mathbb{E}[\omega]= \omega^*\) and he is dogmatic about being of
type \(\hat{\theta}>\theta^*\).
\footnote{The example is illustrated for an overconfident agent but the results are symmetric for a digmatic agent who initially 
places probability one on some $\tilde{\theta}<\theta^*$.} Therefore,
the student's payoff in period \(t\) is given by }

@@

\emph{Under this specification, the myopic optimal effort level is
\(a_t^* = \omega^*\). Nonetheless, because the agent does not know
\(\omega^*\), he will choose \(a_t = \mathbb{E}_t(\omega)\) where the
expectation is taken with respect to the agent's belief at the beginning
of period \(t\). If he does not revise his effort choice for \(k\)
periods, he will receive an average utility of
\((\theta^*+a_t^*)\omega^* - \frac{1}{2}a_t^{*2}\) but he was expecting
an average utility of
\((\hat{\theta}\theta+a_t^*)\omega^* - \frac{1}{2}a_t^{*2}\). In
response, he will apply bayes rule to update his beliefs about
\(\omega\) to get the posterior belief with
\(\mathbb{E}_{t+k}[\omega] = \frac{(\theta^{*} + \omega^{*})\omega^{*}}{\hat{\theta} + \omega^{*}}\)
which is lower than the initial belief. This will cause the agent to
choose a lower effort at \(t+k\). As a result, he will again receive an
average utlitity that is lower than what he expected which will cause
his belief to drift further down. This process will continue until the
average utility equals his expected utility under the dogmatic belief
that assigns probability 1 to \(\hat{\theta}\). At that point, the
stundent will have reached a self-defeating equilibrium and he will
continue to choose sub-optimal effort forever. }

Although the model of a dogmatic modeler is able to rationalize the
prevalence of overconfident (underconfident) beliefs, the assumption
that the agent has a degenerate belief and no mechanism through which he
can update such belief is very restrictive. An alternative approach is
proposed by \citet{Ba2023}. She proposes an extension of the dogmatic
agent who is able to jump from one dogmatic belief to another. By doing
so, the agent might end up being dogmatic and correctly speciefied.

@@

An agent is a \emph{switcher} if they behave as a dogmatic, but is
willing to entertain the possibility that they are of a different type.
In particular, when they start off as a misspecified dogmatic, they are
willing to switch to a different dogmatic belief if the data is
convincing enough. Notice that their beliefs are still degenerate and
assign probability one to a particular type, and zero to all other
types. This means that a bayesian update on \(\theta\) would cause no
change in their beliefs. However, they are willing to entertain two such
beliefs and have a mechanism by which they decide which belief to adopt
at any period \(t\).

@@

\[\frac{p[h^t|\tilde{\theta}]}{p[h^t|\hat{\theta}]} > \alpha\geq 1\]

Where \(h^t\) is the history of outcomes up to time \(t\) and \(\alpha\)
is the switching threshold. By keeping track of the likelihood ratio,
the agent can perform a \emph{Bayesian hypothesis test} and adopt the
Dogmatic belief that best fits the
data.\footnote{In a related problem @Schwarstein2021 proposes a similar updating procedure wich relies on the Bayesian 
hypothesis test.} Notice that if \(\alpha \to \infty\), the behavior of
the switcher will be indistinguishable from that of the Dogmatic
modeler. In this sense, the switcher is a generalization of the dogmatic
type.

By allowing the agent to keep track of the likelihoods and switching to
an alternative type, the switcher can avoid the self-confirming
equilibrium. However, if the prior belief on \(omega\) is sufficiently
tight around a self-defeating equilibrium, the switcher might look
identical to the dogmatic even in a case where \(\alpha\) is not too
large. This happens because under the agent's prior, the likelihood
ratio is unlikely to grow as fast as it is needed to escape the
self-defeating equilibrium. In this cases we say that the misspecified
belief is persistent. Hence, in order to be able to determine which of
these two models provides a better explanation of the observed
behaviors, we must focus on cases in which the prior is diffused enough
and the sequence of realized outcomes is such that the self-defeating
equilibrium would not persist.

\hypertarget{framework-2}{%
\section{Framework 2}\label{framework-2}}

The agent is still of some type \(\theta^* \in \Theta\) and the state is
\(\omega \sim F\). In this case the agent chooses an action \(a\in A\)
and observes a binary outcome that is either a success or a failure.
Denote the outcome by \(o \in {s,f}\). The probability of observing a
success is ingreasing in \(\theta^*\) and in \(\omega\). Whenever the
agent observes a succes, he gets a payoff \(v>0\) and whenever the
outcome is a failure, normalize the payoff to 0. In addition, the
probability of success is such that for each state, there is a unique
optimal action that maximizes the agent's expected payoff.

Two nested theories that have been widely studied fall within this
framework: Fully Bayesian updating and self-serving attribution bias. I
explain each of these classical models of belief updating in what
follows

\hypertarget{the-bayesian}{%
\subsection{The Bayesian}\label{the-bayesian}}

A Bayesian agent simultaneously updates their beliefs about \(\theta\)
and \(\omega\) by using Bayes' rule. The posterior odds at period t
about \(\theta\) after observing an outcome are given by: \[
\frac{p_{t}[\theta_H|\text{outcome}]}{p_{t}[\theta_M|\text{outcome}]} = 
      \frac{p[\text{outcome}|\theta_H]p_{t-1}[\theta_H]}{p[\text{outcome}|\theta_M]p_{t-1}[\theta_M]}
\] and \[
\frac{p_{t}[\theta_M|\text{outcome}]}{p_{t}[\theta_L|\text{outcome}]} = 
      \frac{p[\text{outcome}|\theta_M]p_{t-1}[\theta_M]}{p[\text{outcome}|\theta_L]p_{t-1}[\theta_L]}
\]

Bayesian agents always choose the effort level that maximizes their flow
payoff by taking expectations over their prior beliefs about \(\theta\)
and \(\omega\). Since agents are myopic, even though all the parameters
could be identified with enough variation in choices, although the
Bayesian agent is the closest to a fully rational agent discussed here,
they might not learn their true type. This happens because, by being
myopic, they do not internalize the tradeoff between flow payoff and
learning. This can result in too little experimentation to learn their
true type. An alternative to this approach is given by
\citet{Hestermann2021} and is discussed with the results.

\hypertarget{the-self-serving-bayesian}{%
\subsection{The Self-Serving Bayesian}\label{the-self-serving-bayesian}}

A self-serving bayesian is an agent who uses a biased version of Bayes
rule to update his beliefs.He will update his beliefs about the state
\(\omega\) and his type \(\theta\) by over-attributing successes to a
high value of \(\theta\) and under-estimating the role of higher
\(\omega\). Similarly, he will attribute failure to a low state to a
greter degree than an unbiased agent would. To model the self-serving
attribution bias, I take the approach of \citet{benjamin2019}, where the
posterior is given by:

\[
p_{t}[\theta_H|\text{outcome}] = 
\frac{p[\text{outcome}|\theta]^{c_s^{\theta}\mathbb{I}\{\text{success}\}+c_f^{\theta}\mathbb{I}\{\text{failure}\}}p_{t-1}[\theta]}
{\sum_{\theta'\in\Theta}p[\text{outcome}|\theta']^{c_s^{\theta}\mathbb{I}\{\text{success}\}+c_f^{\theta}\mathbb{I}\{\text{failure}\}}p_{t-1}[\theta']}
\]

@@

\hypertarget{a-unifying-example}{%
\section{A Unifying Example}\label{a-unifying-example}}

The agent can be of one of 3 types:
\(\theta \in \{\theta_L, \theta_M, \theta_H\}\) with
\(\theta_H > \theta_M > \theta_L\). They face an unknown exogenous
success rate \(\omega \in \{\omega_L, \omega_M, \omega_H\}\) with
\(\omega_H>\omega_M>\omega_L\). Each of the values of \(\omega\) is
realized with equal probability. The agent knows the distribution of
\(\omega\) but not its realized value.

Denote the true type by \(\theta^*\) and the true state by \(\omega^*\).
The agent holds some prior belief about \(\theta\)
\footnote{which is potentially misspecified as in the dogmatic and switcher cases discussed above}
and chooses a binary gamble \(e in \{e_L, e_M, e_H\}\). The agent
observes whether the gamble is a success or a failure and gets a payoff
of \(1\) and if it is a success; they get \(0\) otherwise.

The probability of success is increasing in both \(\theta\) and
\(\omega\) and is fully described by the following table:

@@

Conditional on a type, the agent's flow payoff is maximized by choosing
the gamble that matches the state. For example, if the value of
\(\omega\) is \(\omega_H\), the agent's flow payoff is maximized by
choosing \(e_H\) and if the state is \(\omega_L\) the flow payoff is
maximized by choosing gamble \(e_L\), regardless of the value of
\(\theta\). The agent myopically chooses gambles every period to
maximize the flow payoff for \(T<\infty\) periods.

After observing the outcome of each gamble, the agent updates their
beliefs using some procedure and moves on to the next period.

Notice that both \(\theta\) and \(\omega\) can be identified from the
outcomes if enough variation in the effort choices exists. This can be
seen by confirming that there is no pair of \(\theta\) and \(\omega\)
such that the probability of success is the same for all effort choices.
Thus, by changing the effort choice, the agent can learn both their type
and the state if they observe enough outcomes.

In this example, for an agent with a dogmatic belief about their type, a
self-defeating equilibrium is one in which the agent chooses an effort
level that, under the true \(\theta\), yields a frequency of success
that is consistent with the agent's misspecified belief. That is
\(P[\text{sucess}|\theta^*, e^*] = P[\text{sucess}|\hat{\theta}, e^*]\)
where \(e^*\) is the agent's myopic optimal choice.

In the data-generating process described above, there are five such
equilibria. For example, if the agent is of type \(\theta_M\) but
mistakenly believes that he is of type \(\hat{\theta}=\theta_H\) and the
and \(\omega^* = \omega_M\), when the effort chosen is \(e_L\), the
agent will observe a success with 45\% chance. Because the agent
dogmatically believes that their type is high, they will erroneously
conclude that the rate is \(\omega_L\). Under this belief, the optimal
action is \(e_L\) which will continue to generate successes with \(45%
\) probability, further reinforcing the incorrect belief. By doing so,
the agent forgoes the payoff from gamble \(e_M\) which would yield a
success with \(65\%\) chance.

By including self-confirming equilibria, the example captures the forces
from each of the updating mechanisms discussed in the previous section
and allows for direct comparison of all the theories. For realizations
of \((\theta, \omega)\) for which there are self-confirming equilibria,
the dogmatic agent will fall into the trap whereas the switcher will be
able to escape it. Similarly, an agent with self-attribution bias will
update their beliefs differently from an unbiased Bayesian, leading them
to choose different gambles. I exploit such cases in order to test which
model is a better fit for how subjects behave in a laboratory
experiment.
\footnote{because the setting does not match that of @Heidhues2018, there will be situationf for which the theory 
does not provie a prediction. If such cases arise in the lab, they will not be used for the analysis. However, whether a misspecified 
belief persists or not for the switcher, depends highly on the realized history of signals that he gets.}
In what follows I explain the details of how this example was
implemented in the lab.

@@

The experiment consisted of 2 treatments: the \emph{ego-relevant}
condition and the \emph{stereotype} condition. Subjects participated in
only one of the treatments. Treatments were randomly assigned at the
session level. The tasks were identical across treatments, except for
parameter \(\theta\). In the ego-relevant condition \(\theta\) is the
subject's own performance in a quiz, while in the stereotype condition,
it is the performance of a randomly selected subject from another
session.
